Methods

### 2022-08-04 article/intro.Rmd
rev meth,intro MES need disc,abs

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*, and while tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and regenerating degraded functions *[@hubau20]* amid increasing climate stressors *[@anderegg22]* yet insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, as well as land-use history, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]* including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
Overall, while edge effects near degraded habitat edges are often negative, meta-population *[@levins79]* and -community *[@leibold04; @warren15]* theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.

@@

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
**- what is new in this study?**
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade that would suppress less shade-tolerant pioneer tree taxa and allow for more tree diversity closer to the primary forest edge.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

>>>>

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*.
While tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and functionally degraded *[@hubau20]*.
This likely amplifies amid increasing climate stressors *[@anderegg22]* and collective management issues, including insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]*, together with the composition of surrounding forests.
Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
More specifically, timber plantations have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
This could lead to a longer-term dominance of pioneer taxa, referred to as arrested succession *[@]*.
However, arrested succession could be prevented and even typical succession accelerated if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
Additionally, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects near degraded habitat edges are often negative, but influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
Taken together, these ideas suggest that near primary forest, shade-tolerant dispersal may facilitate plantation forest regeneration.
This kind of edge effect may ultimately help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.

@@

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
**- what is new in this study?**
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

### 2022-08-04 article/methods.Rmd
rev meth,intro MES need disc,abs

Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
Given the small size of the last 300 m distance stratum (Fig \@ref(fig:map)c), one census plot was sufficiently representative.
All data were later centered at the stratum level, so resulting trends were not weighted by distance bands with more plots in them. 
The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

@@

The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,243 trees.
<!-- Following common forestry guidelines,  -->
In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.
Taxa were identified by local experts, 
and trait information was gathered from the literature.
Traits including successional stage (early, late) and main dispersal mode (wind, animal, water) were matched to species with *@werden20* data, and wood tissue carbon matched to genera with the GLOWCAD database *[@doraisami22]* given high alignment at species levels *[@chave06]*.

@@

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis, including unidentified taxa, which were grouped together for community analyses and did not affect results.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.

>>>>

Edge effects were studied by dividing the secondary forest area into six discrete 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \@ref(fig:map)), using available GIS software (*ArcGIS 10*, [esri.com](esri.com), and *QGIS 2*, [qgis.org](qgis.org)).
Note that in line with hypotheses about facilitated regeneration, this study most often referred to the "edge" of the regenerating plantation forest as the part closest to the unmanaged intact forest, where trees may be dispersing from, rather than the open service road, which is usually the focus of other studies evaluating effects of forest fragmentation and degradation, rather than restoration.
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
Given the small size of the last 300 m distance stratum (Fig \@ref(fig:map)c), one census plot was sufficiently representative.
All data were later centered at the stratum level, so resulting trends were not weighted by distance bands with more plots in them. 
The total area of the 30 census plots equaled ~1 ha or 5% of the total secondary forest stand area, which is comparable to similar studies *[@onyekwelu16]*.

@@

The diameter of all stems >10 cm wide were recorded in each census plot, totaling over 1,243 trees.
<!-- Following common forestry guidelines,  -->
In cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem. 
Tree height was recorded by measuring distances to both the crown and the stem at chest height *(~2.7 m)* using a rangefinder *(Bushnell, Forestry Suppliers, Inc.)* and triangulating the missing side length.

Taxa were identified by local experts, 
and trait information was gathered from the literature.
Final traits used only included successional stage (early, late) and main dispersal mode (wind, animal, water), which were ultimately matched to species using only the specific dataset made available by *@werden20*.
Otherwise wood tissue carbon was matched to genera with the GLOWCAD database *[@doraisami22]*, given high taxonomic alignment of this trait above the species level *[@chave06]*.

@@

<!-- Correlated variables were grouped into multi-way ANOVAs---namely biomass, wood density, and diameter in one group; and height, stem density, and wood density in a different group.  -->
Non-linear regressions were run using the function *poly()* in the base R package *stats* `r packageVersion("stats")`, 
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models. 
All trees censused were included in each analysis, including unidentified taxa, which were only grouped together for taxonomic analyses and whose exclusion did not affect observed patterns shown in results.
Community analysis and ordination was done as a PERMANOVA with the *adonis()* function from the *vegan* `r packageVersion("vegan")` R package *[@dixon03]*.
Data and code were organized with R packages *here* *[@here]*, *bookdown* *[@bookdown2022]*, and *grateful* *[@grateful]*, and internal pipeline *oir* *[@softwareMedina22oir]*, and are stored at [github.com/nmedina17/osa](github.com/nmedina17/osa) *[@softwareMedina22osa]*.
